---
title: "Roadmap to recovery revealed through the reintroduction of an IUCN Red List species"
author: "Wilson B A, Evans M J, Gordon I J, Pierson J C, Brockett B M, Wimpenny C, Batson W G, Newport J, & Manning A D"
date: "26 January 2023"
output:
  html_document:
    toc: true
    number_sections: true
    toc_depth: 3
    toc_float:
      collapsed: false
    theme: cerulean
    highlight: pygments
editor_options:
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(results='hide', warning=FALSE, message=FALSE)
```

```{r caching_setup}
# Setup whether to use cached or generated results
use_cached_results=TRUE
store_cached_results=TRUE
```

<br />

Reintroductions are powerful tools for tackling biodiversity loss, but resulting populations can be intrinsically small and vulnerable. It is therefore critical to maximise the number of individuals available to contribute to recovery efforts. We investigated whether demographic parameters from a reintroduced population of eastern quolls (*Dasyurus viverrinus*) at [a fenced reserve](https://www.mfgowoodlandexperiment.org.au/) could reveal threats to long-term persistence, inform thresholds for management interventions, and create targets for removing this endangered species from the [IUCN Red List](https://www.iucnredlist.org/species/6296/21947190).

In this analysis, we:

  1. Calculated demographic population estimates using the [package RMark (Laake 2013)](http://cran.salud.gob.sv/web/packages/RMark/RMark.pdf) to interface with the [program MARK](http://www.phidot.org/software/mark/downloads/) 
    - Note that this program needs to be installed for the analysis to run.
  2. Plotted our population viability analysis results (PVA) using the [package vortexR (Pacioni & Mayer 2017)](https://besjournals.onlinelibrary.wiley.com/doi/pdf/10.1111/2041-210X.12786) to collate the results generated with the [program Vortex (Lacy & Pollak 2013)](https://scti.tools/vortex/).

---

# Demography

## Data preparation

Firstly, we need to install and load the required packages. The `pacman` package allows you to do this for multiple packages at once, but to use it we first need to manually install the `pacman` package itself.

```{r, eval=FALSE}
# First, install the pacman package manually if not already installed
#install.packages(pacman)
library(pacman)

# Now install (if not already installed) and load the required packages
pacman::p_load(AICcmodavg, boot, dplyr, ggplot2, ggpubr, effects, emmeans, 
               hydroTSM, lme4, lubridate, mgcv, multcomp, MuMIn, plotrix, 
               plyr, readxl, reshape2, RMark, stringi, stringr, tidyverse, 
               viridis)
```

Now we can read in our eastern quoll capture data, which is available at [GitHub](https://github.com/ecologibel).

```{r}
# Set the working directory to where the current Rmd file is saved 
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

# Read in eastern quoll individual data
ind <- read_excel("eastern quoll CMR data.xlsx", 
                     sheet="individuals") %>%
  dplyr::select(c("quoll_code", "sex", "morph"))

# Read in eastern quoll capture data
cap <- read_excel("eastern quoll CMR data.xlsx", 
                  sheet="captures") %>%
  dplyr::select(c("quoll_code", "capture_date", "capture_time", "session", 
                  "age_estimate", "ssr", "condition", "body_weight", 
                  "number_py", "py_age", "pes_l", "pes_r", "head_length")) %>%
  mutate(session=as.factor(session),
         capture_time=as.factor(capture_time))

# Read in eastern quoll capture session data
sess <- read_excel("eastern quoll CMR data.xlsx", 
                   sheet="sessions") %>%
  dplyr::select(c("year", "session", "session_label", "date", "season"))

# Join individual, capture, and session data
ind_cap <- left_join(ind, cap, by="quoll_code")
ind_cap_sess <- left_join(ind_cap, sess, by="session") %>%
  arrange(capture_date, capture_time) %>%
  mutate(capture_date=as.Date(capture_date, format="%d/%m/%Y"), 
         quoll_code=as.factor(quoll_code), 
         night=stri_sub(session,-1), #library(stringi)
         session=substr(session, 1, nchar(session)-1))
```

### Encounter history

The [program MARK](http://www.phidot.org/software/mark/downloads/) calculates population estimates that account for encounter histories (`ch`) of every individual ever encountered in the population. Encounter histories are represented by a string for each individual, where each number in the string represents a trap session, where each individual is either encountered (1) or not (0). For example:

    ch                freq sex
1   0000000000110000    1   F
2   0001101100000000    1   M
3   0000110111000000    1   F

We generate encounter histories for our individuals here.

```{r}
# Generate session night numbers (1 and 2) for primary and secondary trap nights
ind_cap_sess <- ind_cap_sess %>%
  mutate(session_night=paste0(session,"_", night), 
         sex=factor(sex))

# Since some sessions pass over midnight, take the average date
aggregated_dates <- aggregate(capture_date~session_night, 
                              ind_cap_sess, mean)

# Remove the capture_date column so it can be rewritten in the history below
ind_cap_sess$capture_date <- NULL

# Join capture and date data
ind_cap_sess <- left_join(ind_cap_sess, aggregated_dates, 
                          by="session_night")

# Display session nights and their mean dates
table(ind_cap_sess$night, ind_cap_sess$session)

# Build capture history by joining identities and capture_dates,
history <- data.frame(table("quoll_code"=ind_cap_sess$quoll_code,
                            "capture_date"=ind_cap_sess$capture_date))

# removing duplicate captures within each trap night
history <- acast(history, quoll_code~capture_date, value.var="Freq")
history <- as.data.frame(ifelse(history==0, 0, 1))
history$ch <- as.character(do.call(paste0, history))

# Add the sex covariate to the encounter history dataframe 
sex <- data.frame(unique(ind_cap_sess[,c("quoll_code", "sex")]))
sex <- subset(sex, quoll_code!=0)
rownames(sex) <- sex$quoll_code
sex <- sex[, 2, drop=F]
history <- merge(history, sex, by=0)
history$freq <- 1
encounters_ch <- history[,c("ch","freq","sex")]
```

```{r}
if (store_cached_results) {
  save(ind_cap_sess, file="current_data_quoll.RData")
}
```

### Time intervals

Along with encounter histories, [program MARK](http://www.phidot.org/software/mark/downloads/) accounts for the time interval between trap sessions in its calculation of population estimates. Ideally, there should be negligible time interval between primary (capture) and secondary (recapture) sessions (`time_interval` = 0) so we can assume no immigation or emmigration between the sessions.

```{r}
# Extract dates from the dataset and roughly segregate into months
dates <- as.numeric(unique(ind_cap_sess$capture_date))
dates <- (dates-dates[1])/30
months <- trunc(dates, 0)

# Display months so the reader can manually verify the data
print(months)

# Add months to primary sessions
t2 <- c(0, months)
t1 <- c(months, 0)
time.intervals <- (t1-t2)[-c(length(t2), 1)]

# For sessions that passed over two months (e.g, late April–early May)
# replace their values with 0
time.intervals[15] <- 0
```

### Seasons

Here we generate our chosen covariates for input into the CMR models, including `season`, `session`, `trap effectiveness`, and `weather`.

```{r}
# Gather unique sessions and capture dates
seasons <- unique(ind_cap_sess[,c("session","capture_date")])
seasons <- seasons[!duplicated(seasons$session), ]

# Generate Oceania seasons to use as a covariate later
seasons <- seasons %>%
  mutate(year=format(capture_date, format="%Y"),
         season=time2season(capture_date-365/2, 
                            out.fmt="seasons", type="default")) %>%
  mutate(season=str_replace(season, "autumm", "autumn"), 
         season=str_to_title(season),
         months=trunc((as.numeric(capture_date)-
                         as.numeric(capture_date)[1])/30,0) + 1, 
         capture_date=NULL)

# Manually check that the months are as expected for each session
print(seasons)

# Generate month_factor dataframe for use in the CMR models later
seasons <- data.frame(unique(seasons))
month_factor <- factor(seasons$months)
```

<br />

### Trap effectiveness

When we set traps in the field, they can be triggered and closed by the wind, or taken up by non-target species (in our case, eastern bettongs and brushtail possums). This affects the number of traps available for target animals (in our case, eastern quolls) to be caught in, and can thereby affect our population estimates. We created a "trap effectiveness" covariate to account for this in our models.

```{r}
# Read in trap effectiveness data
trap_effect <- read_excel("eastern quoll CMR data.xlsx", 
                     sheet="trap_effectiveness") %>%
  dplyr::select(c("session", "session_date", "traps_checked", "defective_traps", 
                  "effective_traps", "ssr", "cleanskins", "quolls_minus_ssr",
                  "nontarget_bettong", "nontarget_btp", "nontarget_other")) %>%
  replace_na()

# Trim session and night of extra letters
trap_effect <- trap_effect %>%
  mutate(session=str_trim(session), 
         night=str_sub(session, start= -1), 
         session=substr(session, 1, nchar(session)-1), 
         session_date=as.Date(session_date, format="%d/%m/%Y"))

# Create a df of aggregated, average dates for each trap session
aggregated_dates <- aggregate(session_date~session + night, trap_effect, mean)
trap_effect$session_date <- NULL #not needed

# Sum trap effectiveness statistics by session and night
trap_effect <- trap_effect %>%
  group_by(session, night) %>%
  summarise_all(list(sum=sum))

# Join trap effectiveness with aggregated dates, and order by date
trap_effect <- left_join(trap_effect, aggregated_dates, 
                         by=c("session", "night")) %>%
    arrange(session_date)

# Calculate number of available traps
trap_effect <- data.frame(trap_effect, check.names=FALSE) %>%
  mutate(traps=(traps_checked_sum-
                defective_traps_sum-
                nontarget_btp_sum-
                nontarget_bettong_sum-
                nontarget_other_sum), 
         session_date=as.Date(session_date, format="%d/%m/%Y")) %>%
  arrange(session_date)

# Remove first session so we can cbind() later
trap_effect %$%
{ session == lag(session, n=1) } %>% 
  as.numeric() %>% 
  {.} -> trap_effect$same

# Add a column with sub- and first-session values
trap_effect <- trap_effect %>%
  mutate(same=ifelse(night=="A", "first_session", "sub_session"))
```

#### Weather

```{r}
# Read in weather data from the Australian Buureau of Meterology (product IDCJAC0010)
weather <- read_excel("eastern quoll CMR data.xlsx", 
                     sheet="weather") %>%
  dplyr::select(c("year", "month", "day", "max_temp_c", "min_temp_c", "rain_mm"))

# Generate covariates for weather in the 3 months prior to trapping
weather <- weather %>%
  mutate(max_temp_3mp=rollapply(max_temp_c, 30, mean, 
                                na.rm=TRUE, align="right", fill=NA), 
         min_temp_3mp=rollapply(min_temp_c, 30, mean,
                                na.rm=TRUE, align="right", fill=NA), 
         rain_3mp=rollmeanr(rain_mm, 30, mean, 
                            na.rm=TRUE, align="right", fill=NA), 
         date=as.Date(with(weather, paste(year, month, day, sep="-")), "%Y-%m-%d"), 
         day=NULL, month=NULL, year=NULL)
```

```{r}
# Join trap effectiveness and weather dfs 
trap_effect <- left_join(trap_effect, weather, by=c("session_date"="date"))
trap_effect_c <- trap_effect %>%
  subset(same=="sub_session")

# Label rownames so they will be merged later
rownames(trap_effect) <- seq(from=1, to=nrow(trap_effect), by=1)
rownames(trap_effect_c) <- seq(from=1, to=nrow(trap_effect_c), by=1)
```

## MARK

### Models not including sex

Here we run a capture-mark-recapture (CMR) analysis with a Robust Design closed recruitment model. This model gives apparent survival `Phi` and recruitment `f` for each primary session. It also is 'closed' with no emigration or immigration (like Mulligans Flat). The model produces the following estimates: 

  1. `p` capture rate
  2. `c` recapture rate
  3. `Phi` apparent survival (this always varies by `session` and `time`)
  4. `f` recruitment (this always varies by `session` and `time`)

If we wanted `p` and `c` to vary according to something else, such as bait type (for instance, if we were analysing a dataset with both bettongs and quolls), we would need to repeat the procedure as for adding `season`.

```{r}
# Create processed data 'dp' with encounter history, model, and time intervals
dp=process.data(encounters_ch, model="RDPdfClosed",
                time.intervals=time.intervals)

# Create the design data list 'ddl' wherein we can add model covariate
ddl <- make.design.data(dp)

# For capture probability 'p', add season and trap effectiveness covariates
ddl$p <- left_join(ddl$p, seasons, by=c("session"="months_factor"))
ddl$p <- cbind(ddl$p, trap_effect)

#Manually check the order of the 
print(ddl$p)

ddl$p$min_temp <- as.vector(scale(ddl$p$min_temp))
ddl$p$traps <- as.vector(scale(ddl$p$traps))

ddl$c <- left_join(ddl$c,seasons,by=c("session"="months_factor")) #joins the seasons dataset
ddl$c <- cbind(ddl$c,trap_effect_c) #check the order
ddl$c$min_temp <- as.vector(scale(ddl$c$min_temp))
ddl$c$traps <- as.vector(scale(ddl$c$traps))
```

<br />

### Models where sex is not considered

```{r, eval=!use_cached_results}
setwd("MARK")

# Null hypothesis where capture rate = recapture rate
p <- list(formula=~1, share=T)
f <- list(formula=~time)
Phi <- list(formula=~time)
mod.1=mark(dp,ddl, model.parameters=list(p=p,f=f,Phi=f),
           model.name="mod1-p(.)==c(.)")

# Null hypothesis where capture rate and recapture rate can vary
c <- list(formula=~1)
p <- list(formula=~1)
f <- list(formula=~time)
Phi <- list(formula=~time)
mod.2=mark(dp,ddl, model.parameters=list(p=p,f=f,c=c,Phi=f),
           model.name="mod2-p(.)!=c(.)")

# Wherein capture rate and recapture rate are equal, but population size can vary by session
p <- list(formula=~session, share=TRUE)
f <- list(formula=~time)
Phi <- list(formula=~time)
mod.3=mark(dp,ddl, model.parameters=list(p=p,f=f,Phi=f),
           model.name="mod3-p(a)==c(a)")

c <- list(formula=~session)
p <- list(formula=~session)
f <- list(formula=~time)
Phi <- list(formula=~time)
mod.4=mark(dp,ddl, model.parameters=list(p=p,c=c,f=f,Phi=f),
           model.name="mod4-p(a)!=c(a)")

p <- list(formula=~season, share=T)
f <- list(formula=~time)
Phi <- list(formula=~time)
mod.5=mark(dp,ddl, model.parameters=list(p=p,f=f,Phi=f),
           model.name="mod5-p(s)==c(s)")

c <- list(formula=~season)
p <- list(formula=~season)
f <- list(formula=~time)
Phi <- list(formula=~time)
mod.6=mark(dp,ddl, model.parameters=list(p=p,f=f,c=c,Phi=f),
           model.name="mod6-p(s)!=c(s)")

c <- list(formula=~factor(par.index))
p <- list(formula=~factor(par.index))
f <- list(formula=~time)
Phi <- list(formula=~time)
mod.7=mark(dp,ddl, model.parameters=list(p=p,f=f,c=c,Phi=f),
           model.name="mod7-p(t)!=c(t)") #can't get model to run

c <- list(formula=~traps)
p <- list(formula=~traps)
f <- list(formula=~time)
Phi <- list(formula=~time)
mod.8=mark(dp,ddl, model.parameters=list(p=p,f=f,c=c,Phi=f),
           model.name="mod8-p(traps)!=c(traps)")

c <- list(formula=~min_temp_c)
p <- list(formula=~min_temp_c)
f <- list(formula=~time)
Phi <- list(formula=~time)
mod.9=mark(dp,ddl, model.parameters=list(p=p,f=f,c=c,Phi=f),
            model.name="mod9-p(min)!=c(min)")

c <- list(formula=~max_temp)
p <- list(formula=~max_temp)
f <- list(formula=~time)
Phi <- list(formula=~time)
mod.10=mark(dp,ddl, model.parameters=list(p=p,f=f,c=c,Phi=f),
            model.name="mod10-p(max)!=c(max)")

c <- list(formula=~rain_mm)
p <- list(formula=~rain_mm)
f <- list(formula=~time)
Phi <- list(formula=~time)
mod.11=mark(dp,ddl, model.parameters=list(p=p,f=f,c=c,Phi=f),
            model.name="mod11-p(rain)!=c(rain)")
```

<br />

### Combining the results

Our best model was model 7, where capture (`p`) and recapture probabilities (`c`) varied between primary and secondary sessions (i.e., by trap night).

```{r, eval=!use_cached_results}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

model.table <- collect.models(, adjust=FALSE, table=TRUE) #creates model table

# Display the table
model.table

both.table <- model.table(model.table, model.name=FALSE)

# Select the top model
top.mod.both <- get(both.table$model[1])
plotdat.both <- data.frame(top.mod.both$results$derived$`N Population Size`) #extract for plots
both.results <- top.mod.both$results

if (store_cached_results) {
  write.csv(both.table, file="quoll.both.tableJune2022.csv")
  save(plotdat.both, file="quoll.plotdat.both_June2022.RData")
  save(top.mod.both,file="quoll.both.top.modJune2022.RData") #mod13
  save(both.results, file="quoll.both.results_June2022.RData")
  write.csv(both.results$real,file="quoll.both.results.tableJune2022.csv")
}

```

```{r}
# plotdat.both
load(file="quoll.plotdat.both_June2022.RData")

# top.mod.both
load(file="quoll.both.top.modJune2022.RData")

# both.results
load(file="quoll.both.results_June2022.RData")
```

<br />

### Models with sexes as groups

We will now run the models separately for each sex.

```{r, eval=!use_cached_results}

#first we should remove models from R so we don't get them mixed with these models
mt <- collect.models(table=TRUE, adjust=FALSE)
mod_list <- model.table(mt, model.name=FALSE)$model
rm(list=mod_list) #removes all the models

dp=process.data(encounters_ch, model="RDPdfClosed",
                time.intervals=time.intervals, groups="sex")

# Create the design data list where we can change parameters
ddl=make.design.data(dp)

# Append seasons to the design data list as a covariate for the models
ddl$p <- left_join(ddl$p, seasons, by=c("session"="months_factor"))
ddl$p$night<-ifelse(ddl$p$Time==0,"A","B")

ddl$p <- left_join(ddl$p, trap_effect, by=c("session","night")) #check the order
ddl$p$min_temp <- as.vector(scale(ddl$p$min_temp))
ddl$p$traps <- as.vector(scale(ddl$p$traps))

ddl$c <- left_join(ddl$c,seasons, by=c("session"="months_factor")) #joins the seasons
ddl$c <- left_join(ddl$c, trap_effect_c, by=c("session")) #check the order
ddl$c$min_temp <- as.vector(scale(ddl$c$min_temp))
ddl$c$traps <- as.vector(scale(ddl$c$traps))
```

<br />

#### *Models*

```{r, eval=!use_cached_results}
setwd("MARK")

p <- list(formula=~1)
f <- list(formula=~time:sex)
Phi <- list(formula=~time:sex)
mod.1=mark(dp,ddl, model.parameters=list(p=p,f=f,Phi=f),
           model.name="mod1-p(.)==c(.)(sex)")

c <- list(formula=~1)
p <- list(formula=~1)
f <- list(formula=~time:sex)
Phi <- list(formula=~time:sex)
mod.2=mark(dp,ddl, model.parameters=list(p=p,f=f,c=c,Phi=f),
           model.name="mod2-p(.)!=c(.)(sex)")

p <- list(formula=~session, share=TRUE)
f <- list(formula=~time:sex)
Phi <- list(formula=~time:sex)
mod.3=mark(dp,ddl, model.parameters=list(p=p,f=f,Phi=f),
           model.name="mod3-p(a)==c(a)(sex)")

c <- list(formula=~session)
p <- list(formula=~session)
f <- list(formula=~time:sex)
Phi <- list(formula=~time:sex)
mod.4=mark(dp,ddl, model.parameters=list(p=p,c=c,f=f,Phi=f),
           model.name="mod4-p(a)!=c(a)(sex)")

p <- list(formula=~season, share=TRUE)
f <- list(formula=~time:sex)
Phi <- list(formula=~time:sex)
mod.5=mark(dp,ddl, model.parameters=list(p=p,f=f,Phi=f),
           model.name="mod5-p(s)==c(s)(sex)")

c <- list(formula=~season)
p <- list(formula=~season)
f <- list(formula=~time:sex)
Phi <- list(formula=~time:sex)
mod.6=mark(dp,ddl, model.parameters=list(p=p,f=f,c=c,Phi=f),
           model.name="mod6-p(s)!=c(s)(sex)")

c <- list(formula=~factor(par.index))
p <- list(formula=~factor(par.index))
f <- list(formula=~time:sex)
Phi <- list(formula=~time:sex)
mod.7=mark(dp,ddl, model.parameters=list(p=p,f=f,c=c,Phi=f),
           model.name="mod7-p(t)!=c(t)(sex)")

c <- list(formula=~traps)
p <- list(formula=~traps)
f <- list(formula=~time:sex)
Phi <- list(formula=~time:sex)
mod.8=mark(dp,ddl, model.parameters=list(p=p,f=f,c=c,Phi=f),
            model.name="mod8-p(traps)!=c(traps)(sex)")

c <- list(formula=~min_temp_s)
p <- list(formula=~min_temp_s)
f <- list(formula=~time:sex)
Phi <- list(formula=~time:sex)
mod.9=mark(dp,ddl, model.parameters=list(p=p,f=f,c=c,Phi=f),
            model.name="mod9-p(min)!=c(min)(sex)")

c <- list(formula=~max_temp_c)
p <- list(formula=~max_temp_c)
f <- list(formula=~time:sex)
Phi <- list(formula=~time:sex)
mod.10=mark(dp,ddl, model.parameters=list(p=p,f=f,c=c,Phi=f),
            model.name="mod10-p(max)!=c(max)(sex)")

c <- list(formula=~rain_mm)
p <- list(formula=~rain_mm)
f <- list(formula=~time:sex)
Phi <- list(formula=~time:sex)
mod.11=mark(dp,ddl, model.parameters=list(p=p,f=f,c=c,Phi=f),
            model.name="mod11-p(rain)!=c(rain)(sex)")
```

Collect models:

Our best model was also model 7 where capture probability (`p`) and recapture probability (`c`) vary between primary and secondary sessions (i.e., trap nights). Now we save the results and combine the data for both and sexes for plotting.

```{r, eval=!use_cached_results}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

model.table <- collect.models(, adjust=FALSE, table=TRUE) #creates model table

# Display the table
model.table

sexes.table <- model.table(model.table, model.name=FALSE)
top.mod.sexes <- get(sexes.table$model[1]) #selects the top model
plotdat.sexes <- data.frame(top.mod.sexes$results$derived$`N Population Size`) #extract for plotting
sexes.results <- top.mod.sexes$results

if (store_cached_results) {
  save(top.mod.sexes ,file="quoll.sexes.top.modJune2022.RData")
  save(plotdat.sexes, file="quoll.plotdat.sexes_June2022.RData")
  save(sexes.results, file="quoll.sexes.results_June2022.RData")
  write.csv(sexes.table,file="quoll.sexes.tableJune2022.csv")
  write.csv(sexes.results$real, file="quoll.sexes.results.tableJune2022.csv")
}
```

<br />

## Plots

### Population estimates

Here we load the effects. Note that we need to manually add the sex variable into the dataframe - make sure to check values for females, males, and both.

```{r}
if (use_cached_results) {
  load(file="quoll.plotdat.both_June2022.RData")
  load(file="quoll.plotdat.sexes_June2022.RData")
  load(file="quoll.sexes.top.modJune2022.RData")
  load(file="quoll.sexes.results_June2022.RData")
}

plotdat <- rbind(plotdat.both, plotdat.sexes) #check this is correct
plotdat$session <- paste0(seasons$season, " ", seasons$year) #add season to df
plotdat$session <- fct_inorder(plotdat$session) #library(forcats)
plotdat$group <- c(rep("Both", nrow(plotdat.both)),
                   rep("Females", nrow(plotdat.both)),
                   rep("Males", nrow(plotdat.both))) #update this number

seasons$session <- paste0(seasons$season," ",seasons$year)
plotdat <- left_join(plotdat,seasons, by="session")
plotdat$group <- factor(plotdat$group, 
                        levels=c("Both","Females","Males"))

if (store_cached_results) {
  save(plotdat, file="quoll.estimates_June2022.RData")
  write.csv(plotdat, file="quoll.predictedplotdata_June2022.csv")
}
```

Now we plot those effects:

```{r, warning=FALSE}
if (use_cached_results) {
  load(file="quoll.estimates_June2022.RData")
}

pop <- ggplot(plotdat, aes(x=months, y=estimate, 
                   shape=group, fill=group, group=group, col=group)) + 
  geom_vline(xintercept=seq(8.5, max(plotdat$months), 12), lty=3, col="darkgrey") + 
  geom_errorbar(aes(ymin=lcl, ymax=ucl), width=0)  + 
  geom_line(aes(group=group), lty=2, show.legend=FALSE) +
  geom_point(aes(group=group), cex=4) + 
  annotate(geom="text", x=seq(0, max(plotdat$months), 12), 
           y=60, label=unique(seasons$year), 
           hjust=0, colour="darkgrey", size=3) + 
  theme(axis.line=element_line(colour="black"),
        legend.position=c(0.11, 0.76),
        legend.title=element_blank(),
        legend.text=element_text(size=8),
        legend.key=element_rect(colour="transparent", fill="transparent"),
        panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.border=element_blank(),
        panel.background=element_rect(fill="white"),
        axis.text.y=element_text(angle=0, vjust=0.5, colour="black"),
        axis.text.x=element_text(angle=90, vjust=0.5, hjust=1, colour="black"),
        axis.ticks.x=element_blank(),
        plot.margin=margin(0.1, 0.1, 0.1, 0.1, "cm"),
        strip.text.y=element_blank()) +
  scale_y_continuous(limits=c(0, 60), labels=function(x) formatC(x, width=6)) +
  scale_x_continuous(breaks=unique(as.numeric(as.character(plotdat$months))),
                     labels=unique(plotdat$session),
                     limits=c(-1, max(plotdat$months + 2))) + 
  scale_shape_manual(values=c(21, 22, 24)) +
  scale_colour_manual(values=inferno(3, begin=0.4, end=0.8)) +
  scale_fill_manual(values=inferno(3, begin=0.4, end=0.8)) +
  scale_linetype_manual(values=c(2, 1, 3)) + 
  xlab("Trap session") + 
  ylab("Estimated eastern quoll population size (N)") + 
  labs(shape="session", col="session", fill="session") +
  annotate(geom="text", x=63, y=4, label="a", fontface="bold", size=6)
pop
```

```{r, include=FALSE}
jpeg(file="Figure 2a.jpeg", 
     width=5500, height=3500, units="px", res=800)
pop
dev.off()
```

<br />

### Body weight

We can look at mean body weights for females, males, and both across sessions.

```{r}
if (use_cached_results) {
  # ind_cap_sess
  load(file="current_data_quoll.RData")
}

ind_cap_sess <- left_join(ind_cap_sess, seasons, by="session")
ind_cap_sess$session_label <- paste0(ind_cap_sess$season, 
                                                    " ", 
                                                    ind_cap_sess$year)
ind_cap_sess$sex_word <- ifelse(ind_cap_sess$sex=="M", "Males", "Females")

ind_cap_sess$session_label <- fct_inorder(ind_cap_sess$session_label)
ind_cap_sess$sex=factor(ind_cap_sess$sex)
ind_cap_sess <- subset(ind_cap_sess, ssr=="0")
ind_cap_sess$body_weight_kg <- as.numeric(as.character(ind_cap_sess$body_weight_kg))

jill <- lm(body_weight_kg~session_label,data=data)
bob=allEffects(jill)
effects.b <- as.data.frame(bob[[1]])
effects.b$group <- c("Both")

pw <- lsmeans(jill, pairwise ~ session_label)
ph <- cld(pw, alpha=0.05/nrow(effects.b), Letters=letters, adjust="sidak")
ph <- data.frame(ph)
effects.b <- merge(ph, effects.b, by=c("session_label"))
effects.b$.group <- trimws(effects.b$.group)

load(file="quollconditionJune2022.RData")
data=subset(data,data$sex_word=="Males")
jill <- lm(body_weight_kg~session_label,data=data)
bob=allEffects(jill)
effects.m <- as.data.frame(bob[[1]])
effects.m$group <- c("Males")

pw <- lsmeans(jill, pairwise ~ session_label)
ph <- cld(pw, alpha=0.05/nrow(effects.m), Letters=letters, adjust="sidak")
ph <- data.frame(ph)
effects.m <- merge(ph, effects.m, by=c("session_label"))
effects.m$.group <- trimws(effects.m$.group)

load(file="quollconditionJune2022.RData")
data=subset(data,data$sex_word=="Females")
jill <- lm(body_weight_kg~session_label,data=data)
bob=allEffects(jill)
effects.f <- as.data.frame(bob[[1]])
effects.f$group <- c("Females")

pw <- lsmeans(jill, pairwise ~ session_label)
ph <- cld(pw, alpha=0.05/nrow(effects.f), Letters=letters, adjust="sidak")
ph <- data.frame(ph)
effects.f <- merge(ph, effects.f, by=c("session_label"))
effects.f$.group <- trimws(effects.f$.group)
effects=rbind(effects.b,effects.m,effects.f)
effects$group <- factor(effects$group, levels=c("Both","Females","Males"))

seasons$session_label <- paste0(seasons$season," ",seasons$year)
effects <- left_join(effects, seasons, by="session_label")
write.csv(effects, file="quoll.weights.effectsJune2022.csv")
save(effects, file="quoll.weights.effects.Rdata")
```

<br />

#### Plot

Here we generate a plot of mean body weights for `females`, `males`, and `both` across sessions.

```{r, warning=FALSE}
load(file="quoll.weights.effects.Rdata")

weight <- ggplot(effects, aes(x=months, y=fit, 
                              shape=group, fill=group, col=group), label=group)  + 
  geom_vline(xintercept=seq(8.5, max(effects$months), 12), 
             lty=3, colour="darkgrey") + 
  geom_errorbar(aes(ymin=lower, ymax=upper), 
                width=0, position=position_dodge(2.5))  + 
  geom_line(aes(group=group), lty=2, 
            position=position_dodge(2.5), show.legend=FALSE) + 
  geom_point(aes(group=group), position=position_dodge(2.5), cex=4) + 
  theme(axis.line=element_line(colour="black"),
        legend.position=c(0.67, 0.75),
        legend.title=element_blank(),
        legend.key=element_rect(colour="transparent", fill="transparent"),
        panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.border=element_blank(),
        panel.background=element_rect(fill="white"),
        axis.text.y=element_text(angle=0, vjust=0.5, colour="black"),
        axis.text.x=element_text(angle=90, vjust=0.5, hjust=1, colour="black"),
        axis.ticks.x=element_blank(),
        plot.margin=margin(0.1, 0.1, 0.1, 0.1, "cm"),
        strip.text.y=element_blank()) +
  scale_x_continuous(breaks=unique(as.numeric(as.character(effects$months))),
                     labels=unique(effects$session_label), 
                     limits=c(-1, max(effects$months + 2))) + 
  scale_y_continuous(labels=function(x) formatC(x, width=6)) +
  scale_shape_manual(values=c(21, 22, 24)) +
  scale_colour_manual(values=inferno(3, begin=0.4, end=0.8)) +
  scale_fill_manual(values=inferno(3, begin=0.4, end=0.8)) +
  scale_linetype_manual(values=c(2,1,3)) + 
  xlab("Trap session") + ylab("Mean eastern quoll weight (kg)") +
  labs(shape="group", fill="group", colour="group") +
  annotate(geom="text", x=63, y=0.67, label="b", fontface="bold", size=6)
weight
```

```{r, include=FALSE}
jpeg(file="Figure 2b.jpeg", 
     width=5500, height=3500, units="px", res=800)
weight
dev.off()
```

Here we combine the population and weight estimates into a single publication figure.

```{r, include=FALSE}
pop_pub <- pop + xlab("") +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
  
weight_pub <- weight + theme(legend.position='none')

pop_weight <- ggarrange(pop_pub, weight_pub, ncol=1, nrow=2, widths=c(2, 1.2))
pop_weight

jpeg(file="Figure 2.jpeg", width=4750, height=5500, units="px", res=800)
pop_weight
dev.off()
```

<br />

### Linear regression

Here we fitted a linear model with body `weight` (kg) against estimated population size (`N`), to find a significant negative relationship (*p* = 0.034).

```{r}
pacman::p_load(dplyr, lme4, tidyverse)

N <- read.csv("quoll.predictedplotdata_June2022.csv")
N <- N[with(N, order(group, months)),]
kg <- read.csv("quoll.weights.effectsJune2022.csv")
kg <- kg[with(kg, order(group, months)),]

data <- cbind(N, kg)
data$pop <- data$estimate
data$weight <- data$fit

mod <- lm(weight~pop, data=data)
summary(mod)
anova(mod)
```

<br />

### Density and carrying capacity

Finally, we calculated eastern quoll density by dividing the trapped area (485 ha) by the mean `estimate` number of females and males across all sessions excluding autumn 2017 (where the population was still establishing), resulting in **1 female per 19.57 ha (± 2.84)**, and **1 male per 37.51 ha (± 1.06)**, and a maximum **carrying capacity of 51 adult eastern quolls** (based on the Autumn 2021 session estimate).

```{r}
N <- read.csv("quoll.predictedplotdata_June2022.csv")
N <- subset(N, session!="Autumn 2017")
MULLIGANS_HECTARES = 485
QUOLLS_PER_HECTARE = 12.97

female <- subset(N, group=="Females")
MULLIGANS_HECTARES/mean(female$estimate)
std.error(female$estimate)

male <- subset(N, group=="Males")
MULLIGANS_HECTARES/mean(male$estimate)
std.error(male$estimate)

both <- subset(N, group=="Both")
MULLIGANS_HECTARES/mean(both$estimate)
std.error(both$estimate)

print(QUOLLS_PER_HECTARE * MULLIGANS_HECTARES)

max <- subset(N, group=="Both")
max <- subset(max, session=="Autumn 2021")
MULLIGANS_HECTARES/max$estimate
std.error(max$estimate)
max$estimate
```

<br />

# Viability

The [package vortexR (Pacioni & Mayer 2017)](https://besjournals.onlinelibrary.wiley.com/doi/pdf/10.1111/2041-210X.12786) automates the process of collating, plotting, and analysis (e.g., pairwise scenario comparisons, linear regression) on outputs of population viability analyses (PVAs) conducted with the [program Vortex (Lacy & Pollak 2013)](https://scti.tools/vortex/). First, we:

  1. Create a `pop` congregated database of mean parameter values for each scenario run in the nominated project (in this case, `Eastern quoll`)/
  2. Create a `dot` dataframe that contains dot plots of mean values with standard deviations for specific years (here, 0-100).
    - Were there >1 population, the page would have been divided in rows, and plots for each population would have been on different rows.
  3. Format and order the `census`, `year`, and `scenario` variables.

Note that this package performs best in R versions 4.1.1 and 4.1.2.

```{r}
pacman::p_load(ggplot2, janitor, rJava, rstudioapi, scales, stringr, tidyverse, vortexR) #requires R 4.1.1
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

pop <- collate_dat('Eastern quoll', 1000, save2disk=TRUE,
                   dir_in="C:/Users/belsp/G/Projects/Quoll/Recovery/Vortex/VOutput", 
                   dir_out="Processed data", verbose=FALSE) %>%
  clean_names() %>% na.omit() %>%
  mutate(census=as.numeric(as.character(year)),
         year=as.numeric(as.character(census))/3,
         Scenario=factor(scen_name, 
                         levels=c("No intervention", 
                                  "Harvest (3JF, 3JM)",
                                  "Supplement (1F with 3JF, 3JM)",
                                  "Supplement (1F with 3JF, 3JM) and harvest (2JF, 2JM)",
                                  "Supplement (10F with 30JF, 30JM) and harvest (27JF, 27JM)")))
```

  4. Finally, we generate a plot with parameter values (i.e., `n_all`) on the y-axis and `Scenario` on the x-axis.

```{r}
pva <- ggplot(pop, aes(x=year, y=n_all, col=Scenario, 
                       ymin=n_all - se_n_extant, 
                       ymax=n_all + se_n_extant)) +
  geom_line(size=0.5) +
  theme(axis.line=element_line(colour="black"),
        legend.key=element_rect(colour="transparent", fill="transparent"),
        legend.text=element_text(size=8),
        legend.position="bottom", legend.direction="vertical",
        panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.background=element_rect(fill="white")) +
  scale_color_manual(values=c("#FFDB6D", "#D16103",
                              "#478841", "#81CED4", "#04647E"), 
                              labels=function(x) str_wrap(x, width=50)) +
  guides(color=guide_legend(nrow=3, byrow=TRUE)) +
  xlim(0, 50) +
  xlab("Year") +
  ylab("Simulated eastern quoll population size (N)")
pva
```

Finally, we can export the figure as a .jpeg file.

```{r}
jpeg(file="Figure 3.jpeg", 
     width=5200, height=3850, units="px", res=800)
pva
dev.off()
```

```{r, include=FALSE}
sessionInfo()
```