---
title: "Setup"
author: "Wilson B A"
date: "Aug 7, 2025"
output: 
  html_document:
    toc: true
    number_sections: true
    top_depth: 3
    toc_float:
      collapsed: true
    theme: cerulean
    highlight: pygments
    self_contained: true
editor_options:
  chunk_output_type: console
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(dirname(inputFile), 'setup.html')) })
---

This standalone `setup.Rmd` installs commonly used packages and defines custom objects, functions, and conditional behaviours. Review this file to ensure you're familiar with the setup and its components.

```{r force_eval, eval = TRUE}
# Ensures this Rmd runs, even if the parent Rmd has eval = FALSE
knitr::opts_chunk$set(eval = TRUE)
```

# **Install programs**

To run the demographic population estimation, you will need to install the [program MARK](http://www.phidot.org/software/mark/downloads/) so the [package RMark (Laake 2013)](http://cran.salud.gob.sv/web/packages/RMark/RMark.pdf) can interface with it. 

# **Install packages** 

## Manually install packages

This chunk manually installs packages which can't be installed using the `pacman` loader in the second chunk (e.g., `remotes` for compiling packages from GitHub). 

```{r, eval = FALSE, results = 'hide', warning = FALSE, message = FALSE}
# Manually install `pacman` loader
install.packages(pacman)
```

## Install and load packages

This script installs and loads (1) *project-related* packages, and (2) *commonly used* packages. Replace `beepr` in L42 below with packages required for your analyses.

```{r, eval = FALSE, results = 'hide', message = FALSE, warning = FALSE}
# Install and load project-related packages
pacman::p_load(AICcmodavg, boot, effects, emmeans, hydroTSM, 
               mgcv, multcomp, plotrix, plyr, reshape2, rJava, 
               RMark, rstudioapi, stringi, stringr, vortexR)
```

```{r, results = 'hide', message = FALSE, warning = FALSE}
# Install and load commonly used packages
               # Import and export 
pacman::p_load(openxlsx, readxl, writexl, yaml, 
               # Mapping 
               ggmap, sf, 
               # Plotting 
               corrplot, ggpubr, scales, viridis, 
               # Reporting 
               english, kableExtra, rlang, rprojroot, 
                # Statistical analyses 
               lme4, MuMIn, 
               # Workflow utilities 
               beepr, cli, crayon, glue, here, 
                # Wrangling 
               janitor, tidyverse) 

# Ensure dplyr::select is active to avoid masking by other packages
select <- dplyr::select
```

## Set up `beepr`

Add "`; beep(sound)`" after a line of code to enable a sound notification. This will notify you when long-running code (which you can mark with a ðŸ¢ symbol) finishes evaluating. 

```{r}
# Assign beepr sound (options include coin, complete, facebook, 
# fanfare, mario, ping, ready, shotgun, sword, treasure, and wilhelm)
sound <- "coin"
```

# **Set environment** 

## Conditional behaviours 

Define *conditional behaviours* to control whether downstream code is run or included in the knitted output. 

```{r}
# Assign which behaviours to run when knitting the markdown
knit_options <- list(draft = FALSE, # TRUE runs 'draft' chunks
                     plots = FALSE,  # TRUE runs 'plot' chunks
                     use_cached_results = FALSE, # TRUE uses cached results
                     store_cached_results = FALSE) # TRUE stores cached results
```

### Chunk header condition

If `draft = FALSE`, any chunks with the header option `eval = knit_options$draft` will be skipped during knitting. 

```{r, eval = knit_options$draft}
# This code only runs if draft = TRUE
```

### Conditional code

If you want the chunk to knit but only run part of the code when `draft = TRUE`, wrap it in this `if` block: 

```{r}
# This code only runs if draft = TRUE but the chunk will still knit
if (knit_options$draft) {
  # your code
}
```

## Paths 

```{r}
# Define paths for objects
raw_data       <- "input/raw data.xlsx"
processed_data <- "output/processed data.xlsx"
```

## Settings 

```{r}
# Global setting to deactivate sci notation
options(scipen = 100, digits = 4)
```

# **Set resources** 

## Functions 

### Format text 
Use `format_text()` to parse the format of **bold**, *italic*, and inline `code` formatting from spreadsheets and displays a styled table.

```{r}
# Define function to read a spreadsheet and format it in a table
format_text <- function(path, sheet) 
  {read_excel(path, sheet = sheet) %>%
     # Parse font formatting using named regex replacements
     mutate(across(everything(), ~stringr::str_replace_all(as.character(.x), 
         # Parse **bold** to <strong> 
       c("\\*\\*([^*]+)\\*\\*" = "<strong>\\1</strong>",  
         # Parse *italic* to <em> 
         "\\*([^*]+)\\*"       = "<em>\\1</em>", 
         # Parse `code` to <code> 
         "`([^`]+)`"           = "<code>\\1</code>")))) %>%
     # Optional: wrap column in <code> if not already
     #mutate(Code = paste0("<code>", Code, "</code>")) %>%
     # Print as a clean table
     kbl(align = "l", escape  = FALSE) %>%
     kable_styling(full_width = FALSE, position = "left")}
```

### Generate citation file

Use `generate_citation()` to extract metadata from `analyses.Rmd` and generate a `citation.cff` file that future users can read and export (e.g., via [GitHub](https://docs.github.com/en/repositories/managing-your-repositorys-settings-and-features/customizing-your-repository/about-citation-files)). This function is supported by two helper functions: `extract_metadata()` and `write_cff_from_rmd()`. 

```{r}
# Define main markdown
main_rmd <- "analyses.Rmd" 

# Define function to extract repo metadata
extract_metadata <- function(rmd = main_rmd) {
  if (!requireNamespace("rmarkdown", quietly = TRUE)) 
    stop("Please install the 'rmarkdown' package.")
  
  meta <- rmarkdown::yaml_front_matter(rmd)

  # Evaluate if in inline R format; otherwise return as-is
  date_val <- tryCatch({
    if (grepl("^`r\\s.+`$", meta$date)) {
      expr <- gsub("^`r\\s|`$", "", meta$date)
      as.character(eval(parse(text = expr)))} else {
      as.character(meta$date)}},
    error = function(e) NA_character_)

  repo <- tryCatch(system("git config --get remote.origin.url", 
                          intern = TRUE), error = function(e) "")
  repo_url <- sub("git@github.com:", "https://github.com/", repo)
  repo_url <- sub(".git$", "", repo_url)

  list(title = meta$title, author = meta$author, 
       date = date_val, repo = repo_url, license = "MIT")}

# Define function to write a citation.cff file
write_cff_from_rmd <- function(rmd = main_rmd,
                               cff_path = here::here("metadata/citation.cff")) 
  {if (!requireNamespace("glue", quietly = TRUE)) 
    stop("Please install ", "the 'glue' package.")
  meta <- extract_metadata(rmd)
  dir.create(dirname(cff_path), showWarnings = FALSE, recursive = TRUE)
  if (file.exists(cff_path)) file.remove(cff_path)
  cff <- glue::glue(
    "cff-version: 1.2.0\n",
    "message: \"If you use this code, please cite the following.\"\n",
    "title: \"{meta$title}\"\n",
    "authors:\n",
    "  - name: {meta$author}\n",
    "date-released: {meta$date}\n",
    "repository-code: {meta$repo}\n",
    "license: {meta$license}\n")
  writeLines(cff, cff_path, useBytes = TRUE)
  message(glue::glue("{cff_path} written"))}

# Define function to print the citation details
generate_citation <- function(rmd = main_rmd) {
  if (!requireNamespace("crayon", quietly = TRUE)) 
    stop("Please install the 'crayon' package.")
  if (!requireNamespace("glue", quietly = TRUE)) 
    stop("Please install the 'glue' package.")

  meta <- extract_metadata(rmd)
  bold <- crayon::bold

  # Always show the full date string in parentheses after the author(s)
  author_date <- glue::glue("{meta$author} ({meta$date})")

  citation <- glue::glue(
    "\nIf you use this code, please cite the following:\n\n",
    "{bold(author_date)} {meta$title}. ",
    "Coexistence Conservation Lab, Fenner School of Environment and Society, ",
    "The Australian National University, Canberra.\n",
    "{bold('Available at:')} {meta$repo}\n",
    "{bold('Date released:')} {meta$date}\n",
    "{bold('License:')} {meta$license}\n\n",
    "For more information, visit https://www.coexistenceconservationlab.org/")
  cat(citation)}

# Create citation file 
invisible(write_cff_from_rmd())

# Print citation details
generate_citation()
```